Towards typed repositories of proofs

### 2010-06-18 papers/MIPS2010/abstract.tex
citations ok
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@115 403e655e-980a-47bd-8c60-1ee5348bb6b7

Practical efforts in the formalization of mathematical results have
naturally led to the questions of how to manage large repositories of
proofs, i.e. what is the \emph{daily workflow} of users of a proof
assistant. How does one elaborate a formal proof? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or, the other
way around, how do one rely on existing work to build up new results?
Many of these questions remain largely unanswered, but the tendency
seems to be to adapt existing methods coming from software
development, as illustrated for example by the introduction of
modules, file-based scripts and separate compilation in proof
assistants like \textsf{Coq} ({\cite{CoqDocWeb}}) or
\textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

We believe that these tools are not adapted for the new, demanding
requirements of proof developments. Indeed, whereas compilation of a
program is usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~\cite{lakatos1964proofs}, and we believe that they should be
taken into account in the realization of mathematical assistants.

In fact, although dedicated tools exist to formalize the description
of languages and their metatheory (e.g. \textsf{Twelf},
{\cite{pfenning1999system}}), and substantial formalizations have been
undertaken~[]\cite{lee07}, we still use legacy tools
based on text representation to manage our developments. The general
goal exposed here is to replace this tool chain and make it
language-aware, both on a syntactic side (AST) and on a semantic side
(typing).

@@

The problems of managing a formal mathematical library have been dealt
with in various proof assistant and mathematical repositories. The
HELM project (\cite{asperti2000content}, \cite{asperti2006content})
was an attempt to create a large library of mathematics, importing
\textsf{Coq}'s developments into a searchable and browsable database.
Most ideas from this project were imported into the \textsf{Matita}
proof assistants (\cite{asperti2007hop}), especially a mechanism of
\emph{invalidation and regeneration} to ensure the global consistency
of its library w.r.t changes, with granularity the whole definitions
or proofs and their dependencies. The MBase project
(\cite{kohlhase2001mbase}) attempts at creating a web-based,
distributed mathematical knowledge database putting forward the idea
of \emph{development graph} (\cite{hutter2000management},
\cite{autexier2000towards}) to manage changes in the database,
allowing semantic-based retrieval and object-level dependency
management.

>>>>

Practical efforts in the formalization of mathematical results have
naturally led to the questions of how to manage large repositories of
proofs, i.e. what is the \emph{daily workflow} of users of a proof
assistant. How does one elaborate a formal proof? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or, the other
way around, how do one rely on existing work to build up new results?
Many of these questions remain largely unanswered, but the tendency
seems to be to adapt existing methods coming from software
development, as illustrated for example by the introduction of
modules, file-based scripts and separate compilation in proof
assistants like \textsf{Coq} ({\cite{CoqDocWeb}}) or
\textsf{Matita}~({\cite{AspertiCTZ07}}), the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

We believe that these tools are not adapted for the new, demanding
requirements of proof developments. Indeed, whereas compilation of a
program is usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~(\cite{lakatos1964proofs}), and we believe that they should be
taken into account in the realization of mathematical assistants.

In fact, although dedicated tools exist to formalize the description
of languages and their metatheory (e.g. \textsf{Twelf},
{\cite{pfenning1999system}}), and substantial formalizations have been
undertaken~(\cite{lee07}), we still use legacy tools
based on text representation to manage our developments. The general
goal exposed here is to replace this tool chain and make it
language-aware, both on a syntactic side (AST) and on a semantic side
(typing).

@@

The problems of managing a formal mathematical library have been dealt
with in various proof assistant and mathematical repositories. The
HELM project (\cite{asperti2006content})
was an attempt to create a large library of mathematics, importing
\textsf{Coq}'s developments into a searchable and browsable database.
Most ideas from this project were imported into the \textsf{Matita}
proof assistants (\cite{AspertiCTZ07}), especially a mechanism of
\emph{invalidation and regeneration} to ensure the global consistency
of its library w.r.t changes, with granularity the whole definitions
or proofs and their dependencies. The MBase project
(\cite{kohlhase2001mbase}) attempts at creating a web-based,
distributed mathematical knowledge database putting forward the idea
of \emph{development graph} (\cite{hutter2000management},
\cite{autexier2000towards}) to manage changes in the database,
allowing semantic-based retrieval and object-level dependency
management.
